Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-04-24 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
Edits and notes in conclusion

The locations of food banks and related food services were obtained from the Hamilton Public Library's Food Access Guide^[http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf]. The guide was updated in April of 2021 to indicate any change affected on the services due to the pandemic. This includes modified business hours, a need to make reservations before frequenting, and locations that have completely shut down in consequence. While some food bank services have a specific a target population, such as prioritizing family with young children aged between 0 and 3 or accepting only those providing proof of low-income status through housing and utility statements, all the food bank services indicated below are designed to accommodate those in need of food at zero to low cost. With our focus on food banks and food services that offer free or low-cost meals at particular locations, we first removed services such as Meals on Wheels and other food access services such as food box, community kitchens, student nutrition programs, and shopping and transportation. In addition, two free meal services held on different days at the same location were collapsed into a single service point for the accessibility analysis. Additional details on the operations of individual facilities is not publicly available, and with the changes in operations it proved unfeasible to collect it. For this reason, the analysis to follow is of accessibility to the location of food banks and services, but not to specific services (e.g., breakfasts vs. food boxes).

@@

We used the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}} to download cross-tabulations of: 1) primary mode of travel per trip by income by place of residence; and 2) age by income by place of residence. These data are from the 2016 Survey (the most recent available), and data are geocoded at the level of Traffic Analysis Zones (TAZ) using the most recent zoning system from 2006. Each parcel point is populated with the proportion of trips by three modes of travel: car (as driver or passenger), transit, and walk.

@@

\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the travel time from parcel $i$ to the food bank. In other words, the expected travel time is the weighted sum of travel times to the food bank, with the weights given by the expected modal split in the TAZ.

@@

For the more suburban clusters of zones, the decrease in accessibility is derived from the closure of locations throughout the city reachable by car. In the cluster of central suburban zones for example, low-income households in the outer ring of zones that exhibit medium to high decreases in accessibility within this cluster appear to be largely auto-dependent in their tripmaking, which each each exhibiting between `r round(modes %>% st_drop_geometry() %>% filter(TAZUID %in% c(5143, 5135, 5121, 5110, 5102, 5092, 5063)) %>% summarize(min(p_car)) *100)`-`r round(modes %>% st_drop_geometry() %>% filter(TAZUID %in% c(5143, 5135, 5121, 5110, 5102, 5092, 5063)) %>% summarize(max(p_car)) *100)`\% of their modal split for car trips. This results in the parcels within these zones having a large number of potentially accessible locations in the travel time matrix. But by extension, the change in accessibility over the pre- and during-COVID-19 time periods is affected not only by the closure of service locations proximate to the zones, but also the locations in the central city. The zone with the greatest decrease in accessibility within this cluster (`r round(accessibility_taz_change %>% st_drop_geometry() %>% filter(TAZUID == 5135) %>% pull(acc_d), 4)`) has a high rate of car trips and connects to the most food bank locations in total as well as those that stayed open or closed. The exception to this is the zone in the middle of this cluster, which has a much higher proportion of transit use `r round(modes %>% st_drop_geometry() %>% filter(TAZUID == 5116) %>% summarize(first(p_transit)) *100)`\%. This modal split weighting results in fewer food bank and service locations accessible from the parcels within this zone which, by extension, limits their access to closed locations reachable within 15 minutes in our calculations.

In the cluster to the south-west, the decrease in accessibility is predominately driven by the closure of a high level-of-service Community Meals provider. However, like the more central suburban zones, low-income households within this cluster also between 90% to 100% auto-dependent in their tripmaking in the TTS. The story is similar for the zone located in the north-west that exhibits the greatest decrease in accessibility. Here, low-income households responding to the TTS conducted 100% of their trips by car and, as a result, parcels within this zone have access to the second-highest number of food bank and service locations within 15 minutes. However, this also means zonal accessibilities are greatly affected by the number of closures throughout the city. Finally, in the city's north end and north-east zones, low-income households exhibit a mixture of tripmaking behaviour in the TTS. Households in some zones take transit more often and one zone in particular has 100\% of its trips by walking while in others, car use makes up 100\% of trips. For these zones, the decrease in accessibility tends to be a product of the closure of several inner-city food bank and service locations.

Just as the effects of the closures appear to have been uneven in space, they also seem to have differently impacted various population segments. Table \ref{tab:accessibility-by-age} shows the estimated number of people in each age group by their level of accessibility before and during the pandemic. Here, it is important to note that the quartiles are relative: people in the top 25\% of accessibility still have lower accessibility during the pandemic than before. In reality, every population group is worse off during the pandemic in terms of their accessibility to food banks and services in the City of Hamilton. However, some age groups were affected more. As seen in the table, the distribution within quartiles for adults changed only slightly, and in fact there are fewer adults in the bottom three quartiles as some found themselves in the (deteriorated) top quartile of accessibility during the pandemic. The story is different for seniors, a greater number of whom are now in the third quartile, due to a combination of seniors facing a worse accessibility situation, and just a few moving from the bottom quartile (pre-pandemic) to the third during the pandemic. Those aged 18 and less appear to have fared worse, and the largest change is in the number of children who were in the third quartile before the pandemic found themselves in the bottom accessibility class during the pandemic. These results suggest that, through a combination of the typical modes of transportation of lower income households, and the spatial distribution of the population, the food bank/service locations closed had a differential impact that more greatly affected the youngest among the population in low income households.

@@

>>>>

The locations of food banks and related food services were obtained from the Hamilton Public Library's Food Access Guide^[http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf]. The guide was updated in April of 2021 to indicate any change affected on the services due to the pandemic. This includes modified business hours, a need to make reservations before frequenting, and locations that have completely shut down in consequence. Table 1 defines each service type and the number of locations pre- and during the COVID-19 pandemic. While some food bank services have a specific a target population, such as prioritizing families with young children aged between 0 and 3 or accepting only those providing proof of low-income status through housing and utility statements, all the food bank services indicated below are designed to accommodate those in need of food at zero to low cost. With our focus on food banks and food services that offer free or low-cost meals at particular locations, we first removed services such as Meals on Wheels and other food access services such as food box, community kitchens, student nutrition programs, and shopping and transportation. In addition, two free meal services held on different days at the same location were collapsed into a single service point for the accessibility analysis. Additional details on the operations of individual facilities is not publicly available, and with the changes in operations it proved unfeasible to collect it. For this reason, the analysis to follow is of accessibility to the location of food banks and services, but not to specific services (e.g., breakfasts vs. food boxes).

@@

We used the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}} to download cross-tabulations of: 1) primary mode of travel per trip by income by place of residence; and 2) age by income by place of residence. These data are from the 2016 Survey (the most recent available). The data are geocoded at the level of Traffic Analysis Zones (TAZ) using the most recent zoning system from 2006 and expansion factors are applied to weight the trips . Each parcel point is populated with the proportion of trips by three modes of travel: car (as driver or passenger), transit, and walk.

@@

\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the travel time from parcel $i$ to the food bank. In other words, the expected travel time is the weighted sum of travel times to the food bank, with the weights given by the expected modal split of trips made by low-income households in the TAZ per the TTS data.

@@

For the more suburban clusters of zones, the decrease in accessibility is derived from the closure of locations throughout the city reachable by car. In the cluster of central suburban zones for example, low-income households in the outer ring of zones that exhibit medium to high decreases in accessibility within this cluster appear to be largely auto-dependent in their tripmaking, which each each exhibiting between `r round(modes %>% st_drop_geometry() %>% filter(TAZUID %in% c(5143, 5135, 5121, 5110, 5102, 5092, 5063)) %>% summarize(min(p_car)) *100)`-`r round(modes %>% st_drop_geometry() %>% filter(TAZUID %in% c(5143, 5135, 5121, 5110, 5102, 5092, 5063)) %>% summarize(max(p_car)) *100)`\% of their modal split for car trips. This results in the parcels within these zones having a large number of potentially accessible locations in the travel time matrix. But by extension, the change in accessibility over the pre- and during-COVID-19 time periods is affected not only by the closure of service locations proximate to the zones, but also the locations in the central city. The zone with the greatest decrease in accessibility within this cluster (-0.0009) has a high rate of car trips and connects to the most food bank locations in total as well as those that stayed open or closed. The exception to this is the zone in the middle of this cluster, which has a much higher proportion of transit use `r round(modes %>% st_drop_geometry() %>% filter(TAZUID == 5116) %>% summarize(first(p_transit)) *100)`\%. This modal split weighting results in fewer food bank and service locations accessible from the parcels within this zone which, by extension, limits their access to closed locations reachable within 15 minutes in our calculations.

In the cluster to the south-west, the decrease in accessibility is predominately driven by the closure of a high level-of-service Community Meals provider. However, like the more central suburban zones, low-income households within this cluster also between 90\% to 100\% auto-dependent in their tripmaking in the TTS. The story is similar for the zone located in the north-west that exhibits the greatest decrease in accessibility. Here, low-income households responding to the TTS conducted 100\% of their trips by car and, as a result, parcels within this zone have access to the second-highest number of food bank and service locations within 15 minutes. However, this also means zonal accessibilities are greatly affected by the number of closures throughout the city. Finally, in the city's north end and north-east zones, low-income households exhibit a mixture of tripmaking behaviour in the TTS. Households in some zones take transit more often and one zone in particular has 100\% of its trips by walking while in others, car use makes up 100\% of trips. For these zones, the decrease in accessibility tends to be a product of the closure of several inner-city food bank and service locations.

Just as the effects of the closures appear to have been uneven in space, they also seem to have differently impacted various population segments. Table \ref{tab:accessibility-by-age} shows the estimated number of people in each age group by their level of accessibility before and during the pandemic. Here, it is important to note that the quartiles are relative: people in the top 25\% of accessibility still have lower accessibility during the pandemic than before. In reality, every population group is worse off during the pandemic in terms of their accessibility to food banks and services in the City of Hamilton. However, some age groups were affected more. As seen in the table, the distribution within quartiles for adults changed only slightly, and in fact there are fewer adults in the bottom three quartiles as some found themselves in the (deteriorated) top quartile of accessibility during the pandemic. The story is different for seniors, a greater number of whom are now in the third quartile, due to a combination of seniors facing a worse accessibility situation, and just a few moving from the bottom quartile (pre-pandemic) to the third during the pandemic. Those aged 18 and less appear to have fared worse, and the largest change is in the number of children who were in the third quartile before the pandemic and found themselves in the bottom accessibility class during the pandemic. These results suggest that, through a combination of the typical modes of transportation of lower income households and the spatial distribution of the population, the food bank/service locations closed had a differential impact that more greatly affected the youngest among the population in low income households.

@@

We use data from the census to model low-income households as a proxy for households that are possibly experiencing food insecurity. However, the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known.

While the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown as approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations.

In the absence of information on how food insecure households travel to food banks and related services, we adopted a normative approach that examines accessibility to food banks reachable within 15 minutes based on standards for accessible drinking water...
